---
title: "Example Charts"
author: "M. Drew LaMar"
date: "6/13/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(tidyr)
```

# Basic plots

## Scatter plot (**Basic**)

```{r}
iris %>% 
  ggplot(aes(x = Sepal.Length, y = Sepal.Width, colour = Species)) +
  geom_point()
```

### Stripchart

#### With jitter (**Basic**)

```{r}
iris %>% 
  ggplot(aes(x = Species, y = Sepal.Length)) + 
  geom_point(position = position_jitter(width = 0.1))
```

#### Connecting paired data (**Recipe**)

```{r}
blackbird <- read.csv(url("http://whitlockschluter.zoology.ubc.ca/wp-content/data/chapter12/chap12e2BlackbirdTestosterone.csv")) %>%
  rename(before = logBeforeImplant, after = logAfterImplant) %>%
  gather(before, after, 
         key = "time",
         value = "logAntibody") %>%
  mutate(time = forcats::fct_relevel(time, "before", "after"))
```

```{r}
blackbird %>% 
  ggplot(aes(x = time, y = logAntibody)) + 
  geom_line(aes(group = blackbird)) +
  geom_point(shape = 21, fill = "firebrick", size=3)
```

## Box plot (**Basic**)

```{r}
p <- iris %>% 
  ggplot(aes(x = Species, y = Sepal.Length)) +
  geom_boxplot()
p
```

How to get underlying computed variables?

```{r}
computed <- ggplot_build(p)$data[[1]]
p$layers[[1]]$geom$required_aes
computed
```

## Violin plot (**Basic**)

```{r}
iris %>% 
  ggplot(aes(x = Species, y = Sepal.Length)) +
  geom_violin()
```

## Histogram (**Basic**)

```{r}
iris %>%
  ggplot(aes(x = Sepal.Length)) +
  geom_histogram()
```

## Bar plot (**Basic**)

### Counts

```{r}
diamonds %>%
  ggplot(aes(x = cut)) + 
  geom_bar()
```

Showing hidden assignment of `y` aesthetic:

```{r}
diamonds %>%
  ggplot(aes(x = cut)) + 
  geom_bar(aes(y = stat(count)))
```

### Proportion

```{r}
diamonds %>%
  ggplot(aes(x = cut)) + 
  geom_bar(aes(y = stat(count)/sum(stat(count))))
```

Here's how to do it using the computed `prop` variable.  It is a *groupwise proportion*, however, and by default ggplot adds a group for each level of the categorical variable.  See [this post](https://rstudio-pubs-static.s3.amazonaws.com/291083_2b0374fddc464ed08b4eb16c95d84075.html) that explains this (and a lot more!)

```{r}
diamonds %>%
  ggplot(aes(x = cut)) + 
  geom_bar(aes(y = stat(prop), group = 1))
```

### Stacked barplot (**Basic**)

```{r}
diamonds %>%
  ggplot(aes(x = cut,
             fill = clarity)) + 
  geom_bar()
```

#### Proportional

```{r}
diamonds %>%
  ggplot(aes(x = cut,
             fill = clarity)) + 
  geom_bar(position = position_fill())
```

### Side-by-side barplot (**Basic/Recipe**)

```{r}
p <- diamonds %>%
  ggplot(aes(x = cut,
             fill = clarity)) + 
  geom_bar(position = position_dodge2())
p
```

```{r}
p <- diamonds %>%
  ggplot(aes(x = clarity)) + 
  geom_bar(aes(y = stat(prop), group = cut)) +
  facet_wrap(~cut, scales = 'free_x', nrow=1)
p
```

## Line plot (**Basic**)

```{r}
economics %>%
  ggplot(aes(x = date, y = unemploy)) + 
  geom_line()
```

## Area plot (**Basic**)

```{r}
economics %>%
  ggplot(aes(x = date, y = unemploy)) + 
  geom_area()
```

## Stacked Histograms (**Basic/Recipe**)

```{r}
sticklebackData <- read.csv("http://whitlockschluter.zoology.ubc.ca/wp-content/data/chapter03/chap03e3SticklebackPlates.csv")
sticklebackData$genotype <- factor(sticklebackData$genotype, levels = c("mm","Mm","MM"))
```

```{r}
sticklebackData %>% 
  ggplot(aes(x = plates)) +
  geom_histogram(breaks = seq(0, 70, by=2),
                 fill = "firebrick",
                 color = "black") +
  facet_wrap(~ genotype, ncol = 1)
```

## Mosaic Plots

See https://stackoverflow.com/a/45688044/8663034

```{r}
df <- diamonds %>%
  group_by(cut, clarity) %>%
  summarise(count = n()) %>%
  mutate(cut.count = sum(count),
         prop = count/sum(count)) %>%
  ungroup()

ggplot(df,
       aes(x = cut, y = prop, width = cut.count, fill = clarity)) +
  geom_bar(stat = "identity", position = "fill", colour = "black") +
  # geom_text(aes(label = scales::percent(prop)), position = position_stack(vjust = 0.5)) + # if labels are desired
  facet_grid(~cut, scales = "free_x", space = "free_x") +
  scale_fill_brewer(palette = "RdYlGn") # +
  # theme(panel.spacing.x = unit(0, "npc")) + # if no spacing preferred between bars
  # theme_void() 
```

Another one:

```{r}
toxoplasma <- read.csv(url("http://whitlockschluter.zoology.ubc.ca/wp-content/data/chapter09/chap09e3ToxoplasmaAndAccidents.csv"))

info <- toxoplasma %>% 
  group_by(accident, condition) %>% 
  summarise(count = n()) %>% 
  mutate(accident.count = sum(count), 
         prop = count/sum(count)) %>% 
  ungroup()

ggplot(info, aes(x = accident, y = prop, width = accident.count, fill = condition)) + 
  geom_bar(stat = "identity", position = "fill", colour = "black") + 
  facet_grid(~ accident, scales = "free_x", space = "free_x") + 
  scale_fill_manual(values = c("firebrick", "goldenrod1"))
```

Try to flip it:

```{r}
# Need to flip categoricals to have left mean top
info$condition <- forcats::fct_relevel(info$condition, "uninfected", "infected")

ggplot(info, aes(x = accident, y = prop, width = accident.count, fill = condition)) + 
  geom_bar(stat = "identity", position = "fill", colour = "black") + 
  coord_flip() +
  facet_grid(rows = vars(accident), scales = "free_y", space = "free_y") + 
  scale_fill_manual(values = c("goldenrod1", "firebrick"))
```

# Statistics

## Regression

```{r}
mpg %>%
  ggplot(aes(x = displ, y = hwy, colour = class)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) + 
  theme(legend.position = "none")
```

```{r}
mpg %>%
  ggplot(aes(x = displ, y = hwy, colour = class)) + 
  geom_point() +
  geom_smooth(aes(x = displ, y = hwy), method = "lm", se = FALSE, inherit.aes = FALSE) + 
  theme(legend.position = "none")
```

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth(aes(colour = "loess"), method = "loess", se = FALSE) +
  geom_smooth(aes(colour = "lm"), method = "lm", se = FALSE) + 
  labs(colour = "Method")
```

## ECDF (**Stat**)

```{r}
spiderData <- read.csv("http://whitlockschluter.zoology.ubc.ca/wp-content/data/chapter03/chap03e2SpiderAmputation.csv")
speedBefore <- subset(spiderData, treatment == "before")
```

Base R.

```{r}
plot(ecdf(speedBefore$speed), 
     verticals = TRUE,
     ylab = "Cumulative relative frequency", 
     xlab = "Running speed before amputation (cm/s)")
```

ggplot2

```{r}
speedBefore %>%
  ggplot(aes(x = speed)) + 
  geom_step(stat = "ecdf") + 
  geom_point(stat = "ecdf")
```

## Confidence intervals (**Stat/Recipe**)

```{r}
iris %>% 
  ggplot(aes(x = Species, y = Sepal.Length)) + 
  geom_point(position = position_jitter(width = 0.1)) + 
  geom_errorbar(stat = "summary", 
                fun.data = "mean_cl_normal", 
                fun.args = list(conf.int = 0.95), 
                width = 0.1, 
                position = position_nudge(x = 0.25)) + 
  geom_point(stat = "summary", 
             fun.y = "mean", 
             position = position_nudge(x = 0.25), 
             size=2, 
             color = "red")
```

## Dynamite plots (**Stat/Recipe**)

In front:

```{r}
iris %>%
  ggplot(aes(x = Species, y = Sepal.Length)) +
  geom_bar(stat = "summary",
           fun.y = "mean") +
  geom_errorbar(stat = "summary",
                fun.data = "mean_cl_normal", 
                fun.args = list(conf.int = 0.95), 
                width = 0.1)
```

In back:

```{r}
iris %>%
  ggplot(aes(x = Species, y = Sepal.Length)) +
  geom_errorbar(stat = "summary",
                fun.data = "mean_cl_normal", 
                fun.args = list(conf.int = 0.95), 
                width = 0.1) +
  geom_bar(stat = "summary",
           fun.y = "mean")
```

## Means as lines (**Stat/Recipe**)

```{r}
locustData <- read.csv("http://whitlockschluter.zoology.ubc.ca/wp-content/data/chapter02/chap02f1_2locustSerotonin.csv")
```

Base R.

```{r}
# Stripchart with options
par(bty = "l") # plot x and y axes only, not a complete box
stripchart(serotoninLevel ~ treatmentTime, 
           data = locustData, 
           vertical = TRUE, 
           method = "jitter", 
           pch = 16, 
           col = "firebrick", 
           cex = 1.5, 
           las = 1,
           ylab = "Serotonin (pmoles)", 
           xlab = "Treatment time (hours)",
           ylim = c(0, max(locustData$serotoninLevel)))

# The following command calculates the means in each treatment group (time)
meanSerotonin = tapply(locustData$serotoninLevel, 
                       locustData$treatmentTime, 
                       mean)
# "segments" draws draws lines to indicate the means
segments(x0 = c(1,2,3) - 0.1, 
         y0 = meanSerotonin, 
         x1 = c(1,2,3) + 0.1, 
         y1 = meanSerotonin, lwd = 2)
```

ggplot2

Note that this is going to be more complicated _as we are wanting ggplot2 to do the statistical transformations for us_.  We could go manual as well, which would look like the following:

```{r}
p <- locustData %>%
  ggplot(aes(x = treatmentTime, y = serotoninLevel)) + 
  geom_point(position = position_jitter(width = 0.1, seed = 42),
             color = "firebrick",
             shape = 16,
             size = 3)

summaryData <- locustData %>% 
  group_by(treatmentTime) %>%
  summarize(y = mean(serotoninLevel),
            yend = mean(serotoninLevel),
            x = first(treatmentTime) - 0.1,
            xend = first(treatmentTime) + 0.1)

p + geom_segment(data = summaryData,
                 aes(x = x, 
                     y = y, 
                     xend = xend, 
                     yend = yend, 
                     group = treatmentTime),
                 size = 1) + 
  scale_y_continuous(limits = c(0, NA))
```

```{r}
locustData %>%
  ggplot(aes(x = treatmentTime, y = serotoninLevel)) + 
  geom_point(position = position_jitter(width = 0.1, seed = 42),
             color = "firebrick",
             shape = 16,
             size = 3) +
  geom_errorbarh(stat = "summary",
                 fun.y = "mean",
                 height = 0,
                 aes(xmin = treatmentTime - 0.1,
                     xmax = treatmentTime + 0.1),
                 size = 1) +
  scale_y_continuous(limits = c(0, NA))
```

Another way:

This uses the [ggstance](https://github.com/lionel-/ggstance) package, which gives horizontal versions of geoms, stats, and positions.  This can be better than using `coord_flip`, which applies to all layers.

Note that in order to implement this, the user would need to be able to specify expressions for the values of `xmin` and `xmax`.  Non-standard evaluation could be used here (see below - note that in practice the expressions will be strings).

```{r}
library(ggstance)
one <- quote(treatmentTime - 0.1)
two <- quote(treatmentTime + 0.1)
locustData %>%
  ggplot(aes(x = treatmentTime, y = serotoninLevel)) + 
  geom_point(position = position_jitter(width = 0.1, seed = 42),
             color = "firebrick",
             shape = 16,
             size = 3) +
  geom_linerangeh(stat = "summary",
                  fun.y = "mean",
                  aes(xmin = !!one,
                      xmax = !!two),
                  size = 1) +
  scale_y_continuous(limits = c(0, NA))
```

One more, using `aes_string`:

```{r}
library(ggstance)
one <- "treatmentTime - 0.1"
two <- "treatmentTime + 0.1"
locustData %>%
  ggplot(aes(x = treatmentTime, y = serotoninLevel)) + 
  geom_point(position = position_jitter(width = 0.1, seed = 42),
             color = "firebrick",
             shape = 16,
             size = 3) +
  geom_linerangeh(stat = "summary",
                  fun.y = "mean",
                  aes_string(xmin = one,
                             xmax = two),
                  size = 1) +
  scale_y_continuous(limits = c(0, NA))
```
