---
title: "Geoms and Stats"
author: "M. Drew LaMar"
date: "1/29/2020"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(magrittr)
```

Looks like if aesthetic is specified in base layer, that is the default.
```{r}
iris %>% ggplot(aes(x = Sepal.Length, y = Sepal.Width, size = Species)) + geom_point(stat = "sum")
```
BUT, if it is not specified, the default for the stat kicks in:
```{r}
iris %>% ggplot(aes(x = Sepal.Length, y = Sepal.Width)) + geom_point(stat = "sum")
```
Problem is, how do you specify NULL now???  Before, an empty specification meant NULL, but not anymore.  Could have reset button prompt use for which default they want: base layer (or) stat default.

# Stats

Basic, statistic, 

 * [+] count (basic, 1d)
 * bin_2d (basic, 2d)
 * [+] boxplot (statistic, 1d)
 * contour (2d)
 * [+] sum (basic, 1d)
 * [+] density (basic, 1d)
 * density_2d (basic, 2d)
 * bin_hex (basic, 2d)
 * [+] bin (basic, 1d)
 * qq_line (statistic, 1d)
 * qq (statistic, 1d)
 * quantile (statistic, 1d)
 * [+] smooth (statistic, 1d)
 * ydensity (basic, 1d)
 * sf (basic, 2d)
 * ecdf (statistic, 1d)
 * ellipse (statistic, 2d)
 * function (custom, 1d)
 * [+] identity (basic, 1d, 2d)
 * summary_2d (custom, 2d)
 * summary_hex (custom, 2d)
 * summary_bin (custom, 1d)
 * summary (custom, 1d)
 * unique (manipulate, 1d)
 * sf_coordinates (2d)
 
# Primitives

## Bar

### Stat = count (default) [Bar plot]

```{r}
mpg %>% 
  ggplot(aes(x = class)) + 
  geom_bar()
```

```{r}
mpg %>% 
  ggplot(aes(x = class)) + 
  geom_bar(stat = "count")
```

Computed variables:

 - **count**
 - **prop**
 
Layer data
```{r}
layer_data(mpg %>% 
  ggplot(aes(x = class)) + 
  geom_bar(stat = "count"))
```

Default aesthetics
```{r}
mpg %>% 
  ggplot(aes(x = class)) + 
  geom_bar(stat = "count", aes(y = stat(count)))
```

### Stat = bin [Histogram/Relative frequency dist]

This is essentially a histogram, but with less options (e.g. no binwidth allowed)

```{r}
mpg %>% 
  ggplot(aes(x = displ)) + 
  geom_bar(stat = "bin")
```

Computed variables:

 - **count**: number of points in bin
 - **density**: density of points in bin, scaled to integrate to 1
 - **ncount**: count, scaled to maximum of 1
 - **ndensity**: density, scaled to maximum of 1

```{r}
mpg %>% 
  ggplot(aes(x = displ)) + 
  geom_bar(stat = "bin", bins = 40)
```

All `stat_bin` arguments should work (`binwidth` giving a warning, but probably shouldn't).
```{r}
mpg %>% 
  ggplot(aes(x = displ)) + 
  geom_bar(stat = "bin", 
           bins = 30, 
           closed = "right", 
           pad = TRUE)
```

Default aesthetics.
```{r}
mpg %>% 
  ggplot(aes(x = displ)) + 
  geom_bar(stat = "bin", 
           aes(y = stat(count)),
           bins = 30)
```

Relative frequency distribution (use `stat(density)`)
```{r}
mpg %>% 
  ggplot(aes(x = displ)) + 
  geom_bar(stat = "bin", 
           aes(y = stat(density)),
           bins = 30)
```

Make sure sums to 1.  First, check out layer data.
```{r}
tmp <- mpg %>% 
  ggplot(aes(x = displ)) + 
  geom_bar(stat = "bin", 
           aes(y = stat(count)),
           bins = 30)
ldata <- layer_data(tmp)
str(ldata)
sum(ldata$density*(ldata$xmax[1]-ldata$xmin[1]))
```

### Stat = identity [Bar plot, manual]

```{r}
(means <- iris %>% 
  group_by(Species) %>% 
  summarize(mu = mean(Sepal.Length)))

means %>%
  ggplot(aes(x = Species, y = mu)) + 
  geom_bar(stat = "identity")
```

`geom_col` also does this
```{r}
means %>%
  ggplot(aes(x = Species, y = mu)) + 
  geom_col()
```

Note that you can also use the `stat = summary` to get this automatically.

### Stat = summary [Bar plot, statistics]

```{r}
iris %>%
  ggplot(aes(x = Species, y = Sepal.Length)) + 
  geom_bar(stat = "summary", fun.y = "mean")
```

What does the layer data look like?

```{r}
tmp <- iris %>%
  ggplot(aes(x = Species, y = Sepal.Length)) + 
  geom_bar(stat = "summary", fun.y = "mean")

ldata <- layer_data(tmp)
str(ldata)
```

Ah, makes sense - will have usual aesthetics.  Note that even though the `y` aesthetic for the `geom_bar` layer is not really inheriting the aesthetic from the base layer, and is instead using the calculated `stat(y)` from the layer data, you will get an error if you try and manually specify the `y` aesthetic as follows:
```{r, eval=F}
iris %>%
  ggplot(aes(x = Species, y = Sepal.Length)) + 
  geom_bar(stat = "summary", 
           fun.y = "mean",
           aes(y = stat(y)))
```

Let's try another one (`sum`):
```{r}
iris %>%
  ggplot(aes(x = Species, y = Sepal.Length)) + 
  geom_bar(stat = "summary", fun.y = "sum")
```

What about counting elements using `length`?  Should be equivalent to `stat = count`.
```{r}
mpg %>%
  ggplot(aes(x = class, y = displ)) + 
  geom_bar(stat = "summary", fun.y = "length")
```

Yep!  Only issue is y-axis label.

## Line/Area

Doing these together as they are very similar.

### Stat = identity (default) [Line plot]

```{r}
ggplot(economics, aes(date, unemploy)) + 
  geom_line()
```

```{r}
ggplot(economics, aes(date, unemploy)) + 
  geom_area()
```

### Stat = count

Useful to connect barplot with lines, but need a grouping constant.

```{r}
mpg %>% 
  ggplot(aes(x = class)) + 
  geom_bar(stat = "count") + 
  geom_line(stat = "count", aes(group = 1))
```

Again, y aesthetic is implied:
```{r}
mpg %>% 
  ggplot(aes(x = class)) + 
  geom_bar(stat = "count") + 
  geom_line(stat = "count", aes(y = stat(count), group = 1))
```

```{r}
mpg %>% 
  ggplot(aes(x = class)) + 
  geom_area(stat = "count", aes(y = stat(count), group = 1))
```

### Stat = bin

Same as `stat = count`, draw line for histogram.  In this case, no grouping needed.  Also, I added `pad = TRUE` to match behavior of `geom_freqpoly` (see below).

```{r}
mpg %>% 
  ggplot(aes(x = displ)) + 
  geom_bar(stat = "bin", 
           bins = 30) +
  geom_line(stat = "bin",
            bins = 30,
            pad = TRUE)
```

This is similar to `geom_freqpoly`.  Note that padding happens by default with `geom_freqpoly` and cannot be changed with the `pad` argument.

```{r}
mpg %>% 
  ggplot(aes(x = displ)) + 
  geom_bar(stat = "bin", 
           bins = 30) +
  geom_freqpoly(bins = 30)
```

```{r}
mpg %>% 
  ggplot(aes(x = displ)) + 
  geom_bar(stat = "bin", 
           bins = 30) +
  geom_area(stat = "bin",
            bins = 30,
            pad = TRUE)
```

```{r}
mpg %>% 
  ggplot(aes(x = displ)) + 
  geom_bar(stat = "bin", 
           bins = 30) +
  geom_area(stat = "bin",
            bins = 30,
            pad = TRUE,
            alpha = 0.3,
            colour = "black")
```

### Stat = density

```{r}
diamonds %>%
  ggplot(aes(x = carat)) +
  geom_line(stat = "density")
```

```{r}
diamonds %>%
  ggplot(aes(x = carat)) +
  geom_area(stat = "density")
```

```{r}
diamonds %>%
  ggplot(aes(x = depth, colour = cut)) +
  geom_line(stat = "density") +
  xlim(55, 70)
```

Compare to `geom_density`.  Note that if you use `stat_density`, position must be identity to replicate this plot!!
```{r}
ggplot(diamonds, aes(x = depth, colour = cut)) +
  geom_density() +
  xlim(55, 70)
```

Let's try to replicate this one:
```{r}
ggplot(diamonds, aes(depth, fill = cut, colour = cut)) +
  geom_density(alpha = 0.1) +
  xlim(55, 70)
```

```{r}
ggplot(diamonds, aes(depth, fill = cut, colour = cut)) +
  geom_area(stat = "density", position = "identity", alpha = 0.1) +
  xlim(55, 70)
```

Let's try another one:
```{r}
ggplot(diamonds, aes(x = carat, y = stat(count), colour = cut)) +
  geom_line(stat = "density", position = "stack")
```

One more:
```{r}
ggplot(diamonds, aes(x = carat, y = stat(count), colour = cut)) +
  geom_line(stat = "density", position = "fill")
```

Better with `geom_area`:
```{r}
ggplot(diamonds, aes(x = carat, y = stat(count), fill = cut)) +
  geom_area(stat = "density", position = "fill", colour = "black")
```

### Stat = summary

```{r}
data.frame(
  x = seq(from = 0, to = 4*pi, length.out = 1000)
) %>% 
  ggplot(aes(x = x, y = x)) + 
  geom_line(stat = "summary",
            fun.y = sin)
```

```{r}
data.frame(
  x = seq(from = 0, to = 4*pi, length.out = 1000)
) %>% 
  ggplot(aes(x = x, y = x)) + 
  geom_area(stat = "summary",
            fun.y = sin)
```

```{r}
data.frame(
  x = seq(from = 0, to = 4*pi, length.out = 1000)
) %>% 
  ggplot(aes(x = x, y = x)) + 
  geom_ribbon(stat = "summary",
              fun.ymin = sin,
              fun.ymax = cos)
```

## Points

### Stat = identity (default) [Scatter plot]

```{r}
iris %>%
  ggplot(aes(x = Sepal.Length, y = Sepal.Width)) + 
  geom_point(stat = "identity")
```

### Stat = count

```{r}
mpg %>%
  ggplot(aes(x = cyl)) + 
  geom_point(stat = "count",
             aes(y = stat(count)))
```

```{r}
mpg %>%
  ggplot(aes(x = cyl)) + 
  geom_point(stat = "count",
             aes(y = stat(prop)))
```

### Stat = sum [geom_col]

Should be similar to `geom_count`
```{r}
iris %>%
  ggplot(aes(x = Sepal.Length, y = Sepal.Width)) + 
  geom_point(stat = "sum",
             aes(size = stat(n), weight = 1))
```

```{r}
iris %>%
  ggplot(aes(x = Sepal.Length, y = Sepal.Width)) + 
  geom_count()
```

```{r}
iris %>%
  ggplot(aes(x = Sepal.Length, y = Sepal.Width)) + 
  geom_point(stat = "sum",
             aes(size = stat(prop)))
```

# Non-primitives

## geom_histogram

```{r}
iris %>%
  ggplot(aes(x = Sepal.Length)) + 
  geom_histogram(bins = 40)
```

```{r}
iris %>%
  ggplot(aes(x = Sepal.Length)) + 
  geom_histogram(binwidth = 0.1)
```

```{r}
iris %>%
  ggplot(aes(x = Sepal.Length)) + 
  geom_histogram(breaks = c(4, 5, 6, 7, 8))
```

### Geom = Bar, Stat = Bin

```{r}
iris %>%
  ggplot(aes(x = Sepal.Length)) + 
  geom_bar(stat = "bin", bins = 40)
```

```{r}
iris %>%
  ggplot(aes(x = Sepal.Length)) + 
  geom_bar(stat = "bin", binwidth = 0.1)
```

```{r}
iris %>%
  ggplot(aes(x = Sepal.Length)) + 
  geom_bar(stat = "bin", breaks = c(4, 5, 6, 7, 8))
```

Explicit aesthetic assignment:
```{r}
iris %>%
  ggplot(aes(x = Sepal.Length)) + 
  geom_bar(stat = "bin", 
           aes(y = stat(count)),
           breaks = c(4, 5, 6, 7, 8))
```

## geom_smooth

```{r}
p <- iris %>% 
  ggplot(aes(x = Sepal.Length, y = Sepal.Width))
```

```{r}
p + 
  geom_point() + 
  geom_smooth(method = "lm")
```

### geom_line w/ stat = smooth + geom_ribbon w/ stat = smooth

```{r}
p + 
  geom_point() + 
  geom_ribbon(stat = "smooth", method = "lm", fill = "grey60", alpha = 0.4) +
  geom_line(stat = "smooth", method = "lm", size = 1, colour = "#3366FF")
```

Explicit aesthetic assignments:

```{r}
iris %>% 
  ggplot(aes(x = Sepal.Length, y = Sepal.Width)) + 
  geom_point() + 
  geom_ribbon(stat = "smooth", 
              aes(ymin = stat(ymin), ymax = stat(ymax)),
              method = "lm",  fill = "grey60", alpha = 0.4) +
  geom_line(stat = "smooth", method = "lm", size = 1, colour = "#3366FF")
```

## geom_count

```{r}
p <- iris %>%
  ggplot(aes(x = Sepal.Length, y = Sepal.Width))
```

```{r}
p + 
  geom_count()
```

### geom_point w/ stat = sum

```{r}
p + 
  geom_point(stat = "sum")
```

Explicit aesthetic assignments:

```{r}
p + 
  geom_point(stat = "sum",
             aes(size = stat(n)))
```

## geom_freqpoly

```{r}
p <- ggplot(diamonds, 
       aes(price, colour = cut))
```

```{r}
p +
  geom_freqpoly(binwidth = 500)
```

### geom_line + stat = "bin" + pad = TRUE

```{r}
p +
  geom_line(stat = "bin", binwidth = 500, pad = TRUE)
```

Explicit aesthetic assignment:

```{r}
p +
  geom_line(stat = "bin", 
            aes(y = stat(count)),
            binwidth = 500, pad = TRUE)
```

Extending:

```{r}
p +
  geom_line(stat = "bin", binwidth = 500, pad = TRUE) + 
  geom_point(stat = "bin", binwidth = 500, pad = TRUE)
```

## geom_boxplot

```{r}
p <- ggplot(iris,
            aes(x = Species, y = Sepal.Length))
```

```{r}
p + 
  geom_boxplot()
```

```{r}
# Not sure how to easily do extreme values yet.  Commented is what I would like to do, but it
#   doesn't work as the y aesthetic is defined twice - once to compute stat, and then second time
#   upon assignment.  Maybe this will work with ggplot2 3.3.0??
p + 
  geom_linerange(stat = "boxplot") +
  geom_crossbar(stat = "summary", fun.data = "median_hilow", fun.args = list(conf.int = 0.5), fill = "white", width = 0.8, fatten = 2.0, size = 0.4) # + 
  # geom_point(stat = "boxplot", aes(y = after_stat(outliers)))
```
